Subject-specific curves for time series forecasting of smart meter demand

### 2019-11-02 sscts-paper-1.Rmd
Tidy code. Remove old comments and rename variables appropriately.

# Modelling
n_max_features <- NULL
fcst_test_periods <- 1:96
fcst_test_dates <- dmy("3/1/2017") + 0:355 # exclude final week of year and first day back due to holiday effects
fcst_train_window <- 120
fcst_min_train_days <- 40
building_attributes <- c(
  "basebldngfeedonly",
  "dxsystem",
  "electricelementheating",
  "centraldist"
)
# TODO: This needs to be updated in more than one place (see test-models.R)
#       whenever predictors change. Not good! Define in one place only.
#       Formula for test_models should be created here based on these vars.
# TODO: Rename to pred_vars because lagged demand is also included.
weather_vars <- c(
  "scaled_temperature",
  "scaled_temperature_lag_12",
  "scaled_temperature_lag_24",
  "scaled_temperature_lag_48",
  "scaled_temperature_lag_72",
  "scaled_temperature_lag_96",
  "scaled_temperature_lag_192",
  "scaled_temperature_lag_288",
  "scaled_temperature_max",
  "scaled_temperature_min",
  "scaled_temperature_avg",
  "scaled_wh_lag_96",
  "scaled_wh_lag_192",
  "scaled_wh_lag_672"
)

@@

  # Scale independent variables
  main_df <- main_df %>%
    rename_at(vars(starts_with("temperature")), ~ paste0("scaled_", .)) %>%
    rename_at(vars(starts_with("wh_lag")), ~ paste0("scaled_", .)) %>%
    mutate(
      temperature = scaled_temperature, # keep original temperature for plotting
      wh_lag_96 = scaled_wh_lag_96
    ) %>% # keep original 96 lag demand for plotting
    mutate_at(vars(starts_with("scaled")), scale_dbl) %>% # scale
    select(
      ts, date, period, bid, wh, temperature, wh_lag_96,
      starts_with("scaled"), !!building_attributes
    )

@@

#### Feature selection ========================================================
if (!file.exists(fs_df_file)) {
  cat(
    "Running feature selection for max of", n_max_features,
    "candidate predictors...\n\n"
  )
  start_time <- Sys.time()
  fs_df <- list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(fs_date = map(
      month,
      ~ min(fcst_test_dates[month(fcst_test_dates) == .x])
    )) %>%
    unnest() %>%
    mutate(features = future_map2(
      fs_date,
      period,
      select_features,
      train_window = fcst_train_window,
      min_train_days = fcst_min_train_days,
      n_var_max = n_max_features,
      .progress = TRUE
    ))

@@

fs_df <- fs_df %>%
  mutate(vars = map(features, ~ .x %>%
    # filter(bic == min(bic)) %>%
    filter(loocv == min(loocv)) %>%
    pull(vars) %>%
    unlist()))

@@

```{r fs-alpha, fig.cap="(ref:fs-alpha)", fig.height=7*1.4}
fs_df %>%
  select(period, month, vars) %>%
  unnest() %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(
    selected = tidyr::replace_na(selected, FALSE),
    vars = fct_rev(factor(vars, levels = weather_vars)),
    season = if_else(month %in% month(c(10:12, 1:3), label = TRUE),
      "October-March", "April-September"
    )
  ) %>%
  group_by(season, period, vars) %>%
  summarise(n_selected = sum(selected)) %>%
  ungroup() %>%
  mutate(
    selected = if_else(n_selected > 0, TRUE, FALSE),
    predictor_type = case_when(
      str_detect(vars, "^scaled_temperature") ~ "Temperature",
      str_detect(vars, "^scaled_wh") ~ "Demand",
      TRUE ~ "Other"
    )
  ) %>%
  filter(n_selected > 0) %>% # don't want any fill for 0
  ggplot(aes(
    x = period, y = vars, alpha = n_selected,
    fill = predictor_type
  )) +
  geom_tile() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  labs(
    x = "Period",
    y = "Predictor",
    fill = "Predictor type",
    alpha = "Number of times selected"
  ) +
  facet_wrap(~season, ncol = 1) +
  scale_x_continuous(breaks = seq(8, 96, 8))
```

@@

```{r attribute-changes, fig.cap="(ref:attribute-changes)"}
p_data <- model_df %>%
  select(period, fcst_date, scenarios) %>%
  # filter(fcst_date %in% (ymd("2017-01-09") + 0:4)) %>%
  filter(fcst_date %in% scenario_dates[2:4]) %>%
  unnest() %>%
  filter(bid %in% scenario_bid)
# TODO: Move this to the test-models script
# mutate(scenario = factor(scenario, labels = c("Original building", "Tenant feed included", "Switch to DX system", "Switch to electric element heating", "Switch to centralised distribution"))) %>%

@@

p_df <- model_df %>%
  select(period, fcst_date, scenarios) %>%
  filter(fcst_date %in% scenario_dates) %>%
  unnest() %>%
  filter(
    bid == scenario_bid,
    scenario %in% c("original", "dxsystem")
  ) %>%
  # TODO: Move this to the test-models script
  mutate(scenario = factor(scenario,
    levels = c("original", "dxsystem"),
    labels = scenario_vars
  )) %>%
  select(scenario, ts, pred) %>%
  spread(scenario, pred) %>%
  mutate(
    saving = get(scenario_vars[2]) - get(scenario_vars[1]),
    # TODO: is this saving calc correct? Should I divide by 4?
    saving = cumsum(saving)
  ) %>%
  gather(var, Wh, -ts) %>%
  mutate(
    facet_row = if_else(var %in% scenario_vars, "Energy usage",
      "Cumulative difference"
    ),
    facet_row = factor(facet_row, levels = c(
      "Energy usage",
      "Cumulative difference"
    ))
  )

@@

```{r fs-all-periods-months, fig.cap="(ref:fs-all-periods-months)", fig.height = 7*1.4}
fs_df %>%
  select(period, month, vars) %>%
  unnest() %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(
    selected = tidyr::replace_na(selected, FALSE),
    vars = fct_rev(factor(vars, levels = weather_vars)),
    predictor_type = case_when(
      str_detect(vars, "^scaled_temperature") ~ "Temperature",
      str_detect(vars, "^scaled_wh") ~ "Demand",
      TRUE ~ "Other"
    )
  ) %>%
  filter(selected == TRUE) %>%
  ggplot(aes(x = period, y = vars, fill = predictor_type)) +
  geom_tile() +
  facet_wrap(~month) +
  # scale_fill_manual(values = c("white", "#3d6b78")) +
  labs(
    x = "Period",
    y = "Predictor",
    fill = "Predictor type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(8, 96, 8))
```

>>>>

# Modelling
n_max_features <- NULL
fcst_test_periods <- 1:96
fcst_test_dates <- dmy("3/1/2017") + 0:355 # exclude final week of year and first day back due to holiday effects
fcst_train_window <- 120
fcst_min_train_days <- 40
building_attributes <- c(
  "basebldngfeedonly",
  "dxsystem",
  "electricelementheating",
  "centraldist"
)
predictor_vars <- c(
  "scaled_temperature",
  "scaled_temperature_lag_12",
  "scaled_temperature_lag_24",
  "scaled_temperature_lag_48",
  "scaled_temperature_lag_72",
  "scaled_temperature_lag_96",
  "scaled_temperature_lag_192",
  "scaled_temperature_lag_288",
  "scaled_temperature_max",
  "scaled_temperature_min",
  "scaled_temperature_avg",
  "scaled_wh_lag_96",
  "scaled_wh_lag_192",
  "scaled_wh_lag_672"
)
formula_all <- as.formula(paste("~ bid +", get_terms(predictor_vars)))

@@

  # Scale independent variables
  main_df <- main_df %>%
    rename_at(vars(starts_with("temperature")), ~ paste0("scaled_", .)) %>%
    rename_at(vars(starts_with("wh_lag")), ~ paste0("scaled_", .)) %>%
    mutate(
      temperature = scaled_temperature, # keep original temperature for plotting
      wh_lag_96 = scaled_wh_lag_96      # keep original 96 lag demand for plotting
    ) %>% 
    mutate_at(vars(starts_with("scaled")), scale_dbl) %>% # scale
    select(
      ts, date, period, bid, wh, temperature, wh_lag_96,
      starts_with("scaled"), !!building_attributes
    )

@@

#### Feature selection ========================================================
if (!file.exists(fs_df_file)) {
  cat(
    "Running feature selection for max of", n_max_features,
    "candidate predictors...\n\n"
  )
  start_time <- Sys.time()
  fs_df <- list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(fs_date = map(
      month,
      ~ min(fcst_test_dates[month(fcst_test_dates) == .x])
    )) %>%
    unnest() %>%
    mutate(features = future_map2(
      fs_date,
      period,
      select_features,
      formula = formula_all,
      train_window = fcst_train_window,
      min_train_days = fcst_min_train_days,
      n_var_max = n_max_features,
      .progress = TRUE
    ))

@@

fs_df <- fs_df %>%
  mutate(vars = map(features, ~ .x %>%
    filter(loocv == min(loocv)) %>%
    pull(vars) %>%
    unlist()))

@@

```{r fs-alpha, fig.cap="(ref:fs-alpha)", fig.height=7*1.4}
fs_df %>%
  select(period, month, vars) %>%
  unnest() %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(
    selected = tidyr::replace_na(selected, FALSE),
    vars = fct_rev(factor(vars, levels = predictor_vars)),
    season = if_else(month %in% month(c(10:12, 1:3), label = TRUE),
      "October-March", "April-September"
    )
  ) %>%
  group_by(season, period, vars) %>%
  summarise(n_selected = sum(selected)) %>%
  ungroup() %>%
  mutate(
    selected = if_else(n_selected > 0, TRUE, FALSE),
    predictor_type = case_when(
      str_detect(vars, "^scaled_temperature") ~ "Temperature",
      str_detect(vars, "^scaled_wh") ~ "Demand",
      TRUE ~ "Other"
    )
  ) %>%
  filter(n_selected > 0) %>% # don't want any fill for 0
  ggplot(aes(
    x = period, y = vars, alpha = n_selected,
    fill = predictor_type
  )) +
  geom_tile() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  labs(
    x = "Period",
    y = "Predictor",
    fill = "Predictor type",
    alpha = "Number of times selected"
  ) +
  facet_wrap(~season, ncol = 1) +
  scale_x_continuous(breaks = seq(8, 96, 8))
```

@@

```{r attribute-changes, fig.cap="(ref:attribute-changes)"}
p_data <- model_df %>%
  select(period, fcst_date, scenarios) %>%
  # filter(fcst_date %in% (ymd("2017-01-09") + 0:4)) %>%
  filter(fcst_date %in% scenario_dates[2:4]) %>%
  unnest() %>%
  filter(bid %in% scenario_bid)

@@

p_df <- model_df %>%
  select(period, fcst_date, scenarios) %>%
  filter(fcst_date %in% scenario_dates) %>%
  unnest() %>%
  filter(
    bid == scenario_bid,
    scenario %in% c("original", "dxsystem")
  ) %>%
  mutate(scenario = factor(scenario,
    levels = c("original", "dxsystem"),
    labels = scenario_vars
  )) %>%
  select(scenario, ts, pred) %>%
  spread(scenario, pred) %>%
  mutate(
    saving = get(scenario_vars[2]) - get(scenario_vars[1]),
    saving = cumsum(saving)
  ) %>%
  gather(var, Wh, -ts) %>%
  mutate(
    facet_row = if_else(var %in% scenario_vars, "Energy usage",
      "Cumulative difference"
    ),
    facet_row = factor(facet_row, levels = c(
      "Energy usage",
      "Cumulative difference"
    ))
  )

@@

```{r fs-all-periods-months, fig.cap="(ref:fs-all-periods-months)", fig.height = 7*1.4}
fs_df %>%
  select(period, month, vars) %>%
  unnest() %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(
    selected = tidyr::replace_na(selected, FALSE),
    vars = fct_rev(factor(vars, levels = predictor_vars)),
    predictor_type = case_when(
      str_detect(vars, "^scaled_temperature") ~ "Temperature",
      str_detect(vars, "^scaled_wh") ~ "Demand",
      TRUE ~ "Other"
    )
  ) %>%
  filter(selected == TRUE) %>%
  ggplot(aes(x = period, y = vars, fill = predictor_type)) +
  geom_tile() +
  facet_wrap(~month) +
  # scale_fill_manual(values = c("white", "#3d6b78")) +
  labs(
    x = "Period",
    y = "Predictor",
    fill = "Predictor type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(8, 96, 8))
```
